---
title: "Advanced Guardrail Patterns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Guardrail Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What This Vignette Covers

`vignette("secureguard")` introduces the defense layers and built-in
guardrails. This vignette covers building custom guardrails for
domain-specific threats, composing guardrails with pass/fail logic,
assembling pipelines, and wiring everything into securer for sandboxed
execution.

## How a Guardrail Pipeline Works

The following diagram shows the flow of data through a `secure_pipeline()`,
the recommended way to wire guardrails into an agent loop:

```
         User prompt
              |
              v
    +--------------------+
    |  check_input()     |    Input guardrails:
    |  - injection       |    prompt injection, topic scope, PII
    |  - topic scope     |
    |  - input PII       |
    +--------------------+
              |
         Pass? ----No----> Return failure (stage: input)
              |
             Yes
              |
              v
         LLM generates code
              |
              v
    +--------------------+
    |  check_code()      |    Code guardrails:
    |  - AST analysis    |    blocked functions, complexity,
    |  - complexity      |    dependencies, data flow
    |  - dependencies    |
    |  - data flow       |
    +--------------------+
              |
         Pass? ----No----> Return failure (stage: code)
              |
             Yes
              |
              v
      Execute in sandbox
        (securer)
              |
              v
    +--------------------+
    |  check_output()    |    Output guardrails:
    |  - PII             |    PII blocking, secret redaction,
    |  - secrets         |    size limits
    |  - size            |
    +--------------------+
              |
         Pass? ----No----> Return failure (stage: output)
              |
             Yes
              |
              v
      Return result to user
      (possibly redacted)
```

Each stage short-circuits on failure: if input guardrails reject the prompt,
the LLM never sees it. If code guardrails reject the generated code, it never
executes. This minimizes both risk and wasted computation.

## Creating Custom Guardrails

The built-in guardrails cover common threats: prompt injection, dangerous
function calls, PII leakage, and secret exposure. But every application has
domain-specific risks that generic guardrails will not catch. An agent that
generates SQL needs SQL injection detection. A healthcare application needs
HIPAA-specific PII patterns. A financial tool needs checks for account numbers
and routing numbers.

Every guardrail, built-in or custom, is an S3 object of class `secureguard`
with four properties: `name`, `type`, `check_fn`, and `description`. The
`new_guardrail()` constructor validates these and returns a guardrail you can
use with `run_guardrail()`, `compose_guardrails()`, and `secure_pipeline()`.
Custom guardrails work with built-in ones because they share the same
interface.

### A SQL Injection Detector

When an LLM generates SQL queries, it may produce syntactically valid but
semantically dangerous output, especially if the user's prompt contains
adversarial patterns. A guardrail can catch these patterns before any query
reaches a database.

```{r sql-injection-guard}
library(secureguard)

guard_sql_injection <- function() {
  sql_patterns <- c(
    "(?i)\\b(?:UNION\\s+SELECT|DROP\\s+TABLE|DELETE\\s+FROM)\\b",
    "(?i)\\b(?:INSERT\\s+INTO|UPDATE\\s+.+\\s+SET)\\b.*?;\\s*--",
    "(?i)'\\s*(?:OR|AND)\\s+['\"]?\\d['\"]?\\s*=\\s*['\"]?\\d",
    "(?i)(?:--|#|/\\*).*(?:SELECT|DROP|INSERT|UPDATE|DELETE)"
  )

  check_fn <- function(x) {
    hits <- vapply(sql_patterns, function(pat) {
      grepl(pat, x, perl = TRUE)
    }, logical(1))

    if (any(hits)) {
      guardrail_result(
        pass = FALSE,
        reason = "Potential SQL injection detected",
        details = list(
          matched_patterns = which(hits)
        )
      )
    } else {
      guardrail_result(pass = TRUE)
    }
  }

  new_guardrail(
    name = "sql_injection",
    type = "input",
    check_fn = check_fn,
    description = "Detects common SQL injection patterns"
  )
}
```

Now use it like any built-in guardrail:

```{r sql-injection-run}
g <- guard_sql_injection()
g

# Safe query
run_guardrail(g, "SELECT name FROM users WHERE id = 42")

# Injection attempt
run_guardrail(g, "SELECT * FROM users WHERE id = 1; DROP TABLE users; --")
```

### A Code Length Limiter

Custom guardrails of type `"code"` work exactly the same way. Here is one that
limits the number of lines in LLM-generated code:

```{r code-length-guard}
guard_code_length <- function(max_lines = 100L) {
  check_fn <- function(code) {
    n_lines <- length(strsplit(code, "\n", fixed = TRUE)[[1L]])
    if (n_lines > max_lines) {
      guardrail_result(
        pass = FALSE,
        reason = sprintf("Code has %d lines (max %d)", n_lines, max_lines),
        details = list(n_lines = n_lines, max_lines = max_lines)
      )
    } else {
      guardrail_result(pass = TRUE, details = list(n_lines = n_lines))
    }
  }

  new_guardrail(
    name = "code_length",
    type = "code",
    check_fn = check_fn,
    description = sprintf("Limits code to %d lines", max_lines)
  )
}

g_len <- guard_code_length(max_lines = 5)
run_guardrail(g_len, "x <- 1\ny <- 2\nz <- x + y")

long_code <- paste(sprintf("x%d <- %d", 1:10, 1:10), collapse = "\n")
run_guardrail(g_len, long_code)
```

### Anatomy of a check_fn

Every `check_fn` must:

1. Accept a single argument (the text or object to check).
2. Return a `guardrail_result()` with at minimum `pass = TRUE` or `pass = FALSE`.
3. Optionally include `reason` (why it failed), `warnings` (advisory notes),
   and `details` (a named list of metadata).

The `@` operator accesses properties on the result:

```{r result-properties}
result <- run_guardrail(guard_code_analysis(), "system('ls')")
result@pass
result@reason
result@details
```

## Composing Guardrails

In practice, you almost always want to run multiple guardrails together:
checking for dangerous functions and excessive complexity, or detecting both
prompt injection and off-topic prompts. secureguard provides two ways to
combine them.

### compose_guardrails(): Same-Type Composition

`compose_guardrails()` merges multiple guardrails of the **same type** into a
single composite guardrail. The result is itself a guardrail, so you
can pass it to `run_guardrail()`, nest it inside another composition, or use
it in a pipeline. Bundle all your code checks into a single "strict code"
guardrail to treat them as one unit.

```{r compose-all-mode}
# Compose three code guardrails -- ALL must pass (default)
strict_code <- compose_guardrails(
  guard_code_analysis(),
  guard_code_complexity(max_ast_depth = 10, max_calls = 50),
  guard_code_dependencies(allowed_packages = c("dplyr", "ggplot2"))
)

strict_code

# Clean code passes all three
run_guardrail(strict_code, "dplyr::filter(mtcars, cyl == 4)")

# system() fails code analysis
run_guardrail(strict_code, "system('whoami')")

# processx fails dependency check
run_guardrail(strict_code, "processx::run('ls')")
```

### mode = "any": At Least One Must Pass

The default `mode = "all"` is the right choice for security checks: all
guards must pass. But sometimes you need the opposite logic: an allowlist
where the input is acceptable if it matches **any** of several categories.
With `mode = "any"`, the composite passes if at least one child guardrail
passes:

```{r compose-any-mode}
# Accept prompts about either statistics OR machine learning
topic_guard <- compose_guardrails(
  guard_topic_scope(allowed_topics = c("statistics", "regression", "t-test")),
  guard_topic_scope(allowed_topics = c("machine learning", "neural network")),
  mode = "any"
)

run_guardrail(topic_guard, "How do I run a t-test in R?")
run_guardrail(topic_guard, "Explain neural network backpropagation")
run_guardrail(topic_guard, "What is the weather today?")
```

### check_all(): Run a List and Collect Results

Sometimes you need individual results from each guardrail rather than a single
composite result. `check_all()` runs a list of guardrails and returns a
summary:

```{r check-all}
guards <- list(
  guard_code_analysis(),
  guard_code_complexity(max_ast_depth = 10),
  guard_code_dataflow()
)

result <- check_all(guards, "x <- mean(1:10)")
result$pass
length(result$results)  # one per guardrail

# Inspect individual results
vapply(result$results, function(r) r@pass, logical(1))
```

When a check fails, `check_all()` collects all failure reasons:

```{r check-all-fail}
result <- check_all(guards, "Sys.getenv('SECRET_KEY')")
result$pass
result$reasons
```

### When to Use compose_guardrails() vs check_all()

Both functions combine multiple guardrails, but they serve different purposes
and return different types:

**Use `compose_guardrails()`** when you want a single guardrail object that
you can pass to `run_guardrail()`, nest inside another `compose_guardrails()`,
or use in a `secure_pipeline()`. The composed guardrail behaves as one unit:
you get a single pass/fail result. This is the right choice when you are
building reusable guardrail configurations (e.g., a "strict code" composite)
that you want to treat as a single check.

**Use `check_all()`** when you need diagnostic detail. It returns individual
results for each guardrail in the list, so you can report exactly which checks
failed and why. This is useful in logging, debugging, and user-facing error
messages where "code guardrail failed" is less helpful than "blocked function
`system()` detected by code_analysis; exceeded max AST depth of 10 per
code_complexity."

In practice, many applications use both: `compose_guardrails()` to build
reusable guardrail groups, and `check_all()` at the top level to get per-group
diagnostics.

## Building Pipelines with secure_pipeline()

Individual guardrails and compositions are useful for targeted checks, but a
production agent needs all three defense layers working together. A pipeline
bundles guardrails for input, code, and output into one object with methods
for each stage. You define your security policy once and apply it to every
agent turn.

### Defining a Pipeline

```{r pipeline-create}
pipeline <- secure_pipeline(
  input_guardrails = list(
    guard_prompt_injection(sensitivity = "high"),
    guard_input_pii(),
    guard_topic_scope(allowed_topics = c("statistics", "data analysis", "R"))
  ),
  code_guardrails = list(
    guard_code_analysis(),
    guard_code_complexity(max_ast_depth = 15, max_calls = 100),
    guard_code_dependencies(allowed_packages = c("dplyr", "ggplot2", "tidyr")),
    guard_code_dataflow(block_network = TRUE, block_file_write = TRUE)
  ),
  output_guardrails = list(
    guard_output_pii(),
    guard_output_secrets(action = "redact"),
    guard_output_size(max_chars = 10000, max_lines = 200)
  )
)
```

### Running Each Stage

```{r pipeline-input}
# Stage 1: validate user input
input_result <- pipeline$check_input("Calculate the mean and sd of mtcars$mpg")
input_result$pass
```

```{r pipeline-code}
# Stage 2: validate LLM-generated code
code_result <- pipeline$check_code("
  library(dplyr)
  mtcars %>%
    summarise(mean_mpg = mean(mpg), sd_mpg = sd(mpg))
")
code_result$pass
```

```{r pipeline-output}
# Stage 3: filter execution output
output_result <- pipeline$check_output("mean_mpg = 20.09, sd_mpg = 6.03")
output_result$pass
output_result$result  # possibly redacted text
```

### Pipeline in an Agent Loop

Call the three `check_*` methods in sequence inside your agent loop. Each
stage short-circuits on failure: if `check_input()` rejects the prompt, you
skip the LLM call entirely. If `check_code()` rejects the generated code,
you skip execution. Here is the complete pattern:

```{r agent-loop, eval = FALSE}
process_turn <- function(pipeline, user_prompt, llm_fn, execute_fn) {
  # 1. Input guardrails

  input_check <- pipeline$check_input(user_prompt)
  if (!input_check$pass) {
    return(list(
      success = FALSE,
      stage = "input",
      reasons = input_check$reasons
    ))
  }

  # 2. LLM generates code
  code <- llm_fn(user_prompt)

  # 3. Code guardrails
  code_check <- pipeline$check_code(code)
  if (!code_check$pass) {
    return(list(
      success = FALSE,
      stage = "code",
      reasons = code_check$reasons
    ))
  }

  # 4. Execute in sandbox
  result <- execute_fn(code)

  # 5. Output guardrails
  output_check <- pipeline$check_output(result)
  if (!output_check$pass) {
    return(list(
      success = FALSE,
      stage = "output",
      reasons = output_check$reasons
    ))
  }

  list(success = TRUE, result = output_check$result)
}
```

## Mixing Custom and Built-In Guardrails

Custom and built-in guardrails share the same interface. You can mix them in
`compose_guardrails()`, `check_all()`, and `secure_pipeline()`. There is no
registration step or plugin system; any `secureguard` object works
everywhere:

```{r mix-custom}
# The SQL injection guard from earlier alongside built-in input guards
input_guards <- compose_guardrails(
  guard_prompt_injection(),
  guard_input_pii(),
  guard_sql_injection()
)

run_guardrail(input_guards, "Please help me write a SELECT query")
run_guardrail(input_guards, "' OR 1=1 --")
```

Similarly for code guardrails:

```{r mix-custom-code}
# Custom length guard composed with built-in code guards
code_guards <- compose_guardrails(
  guard_code_analysis(),
  guard_code_complexity(max_ast_depth = 10),
  guard_code_length(max_lines = 50)
)

run_guardrail(code_guards, "x <- mean(1:10)")
```

## Integration with securer

secureguard analyzes code and outputs to decide whether they are safe.
[securer](https://github.com/ian-flores/securer) provides OS-level
sandboxing that limits what the code **can** do, regardless of what it tries.
secureguard catches known-dangerous patterns before execution; securer
contains unknown threats at the operating system level.

securer is a suggested dependency; all of the patterns above work without
it. The integration adds two things: pre-execution hooks and output guarding
after execution.

### Pre-Execute Hooks

`as_pre_execute_hook()` converts code guardrails into a function that securer
calls before executing each code snippet. It returns `TRUE` to allow execution
or `FALSE` to block it.

```{r pre-execute-hook, eval = FALSE}
library(securer)
library(secureguard)

hook <- as_pre_execute_hook(
  guard_code_analysis(),
  guard_code_complexity(max_ast_depth = 15),
  guard_code_dataflow()
)

sess <- SecureSession$new(pre_execute_hook = hook)
sess$execute("mean(1:10)")        # allowed
sess$execute("system('whoami')")  # blocked by code_analysis
sess$execute("Sys.getenv('KEY')") # blocked by dataflow
sess$close()
```

### Post-Execute Output Guarding

`guard_output()` runs output guardrails on execution results. Guardrails with
`action = "redact"` transform the output rather than blocking it:

```{r guard-output, eval = FALSE}
result <- sess$execute("paste('My API key is', 'AKIAIOSFODNN7EXAMPLE')")

checked <- guard_output(
  result,
  guard_output_pii(),
  guard_output_secrets(action = "redact")
)

if (checked$pass) {
  # Return the (possibly redacted) result to the user
  checked$result
} else {
  paste("Blocked:", paste(checked$reasons, collapse = "; "))
}
```

### Pipeline Hook

A pipeline can produce a pre-execute hook from its code guardrails:

```{r pipeline-hook, eval = FALSE}
pipeline <- secure_pipeline(
  input_guardrails = list(guard_prompt_injection()),
  code_guardrails = list(
    guard_code_analysis(),
    guard_code_dataflow()
  ),
  output_guardrails = list(
    guard_output_secrets(action = "redact")
  )
)

sess <- SecureSession$new(
  pre_execute_hook = pipeline$as_pre_execute_hook()
)

# The session now has code guardrails enforced automatically.
# Input and output guardrails are checked manually:
input_check <- pipeline$check_input(user_prompt)
# ... LLM generates code, session executes it ...
output_check <- pipeline$check_output(execution_result)

sess$close()
```

## Advanced Composition Patterns

The patterns above apply the same guardrails to every request. In practice,
you often need to vary strictness based on context: who the user is, where
the request came from, and what level of trust is appropriate.

### Layered Sensitivity

A public-facing chatbot is exposed to adversarial users and needs
high-sensitivity injection detection and tight topic scoping. An internal
analytics tool used by trusted data scientists can use lower sensitivity to
avoid false positives on legitimate analytical prompts:

```{r layered-sensitivity}
# Public-facing: high sensitivity, strict topic scoping
public_guards <- compose_guardrails(
  guard_prompt_injection(sensitivity = "high"),
  guard_input_pii(),
  guard_topic_scope(allowed_topics = c("data analysis", "statistics"))
)

# Internal tool: lower sensitivity, broader topics
internal_guards <- compose_guardrails(
  guard_prompt_injection(sensitivity = "low"),
  guard_input_pii()
)

run_guardrail(
  public_guards,
  "Continue from where we left off with the regression"
)

run_guardrail(
  internal_guards,
  "Continue from where we left off with the regression"
)
```

### Graduated Code Restrictions

You can do the same with code guardrails. A trusted internal user running
vetted analysis scripts needs fewer restrictions than an untrusted external
user whose prompts generate arbitrary code:

```{r graduated-code}
# Trusted context: only block the most dangerous operations
trusted_code <- compose_guardrails(
  guard_code_analysis(blocked_functions = c("system", "system2", "shell")),
  guard_code_dataflow(
    block_env_access = TRUE,
    block_network = FALSE,
    block_file_write = FALSE
  )
)

# Untrusted context: strict lockdown
untrusted_code <- compose_guardrails(
  guard_code_analysis(),
  guard_code_complexity(max_ast_depth = 10, max_calls = 30),
  guard_code_dependencies(allowed_packages = c("dplyr", "ggplot2")),
  guard_code_dataflow(
    block_env_access = TRUE,
    block_network = TRUE,
    block_file_write = TRUE,
    block_file_read = TRUE
  )
)

# The same code may pass in trusted but fail in untrusted
code <- "readLines('data.csv')"
run_guardrail(trusted_code, code)
run_guardrail(untrusted_code, code)
```

### Redact vs Block Decision

PII like social security numbers or patient records should block the entire
response; partial disclosure is still a privacy violation. API keys and
tokens can often be redacted in place, keeping the useful parts of the
response while replacing the sensitive value. Output guardrails support
three actions (`"block"`, `"redact"`, `"warn"`) for this:

```{r redact-vs-block}
# PII blocks the output entirely
# Secrets get redacted so the response is still useful
pipeline <- secure_pipeline(
  output_guardrails = list(
    guard_output_pii(),                         # blocks on PII
    guard_output_secrets(action = "redact"),     # redacts secrets
    guard_output_size(max_chars = 5000)          # blocks oversized output
  )
)

# Secrets are redacted, not blocked
result <- pipeline$check_output("API key: AKIAIOSFODNN7EXAMPLE, data looks good")
result$pass
result$result

# PII causes a block
result <- pipeline$check_output("Patient SSN: 123-45-6789")
result$pass
result$reasons
```

## Summary

| Pattern | Function | Use Case |
|---------|----------|----------|
| Custom guardrail | `new_guardrail()` | Domain-specific checks |
| Same-type composition | `compose_guardrails()` | Merge guards into one reusable unit |
| Batch check | `check_all()` | Individual results per guard (diagnostics) |
| Full pipeline | `secure_pipeline()` | Three-layer defense for production |
| Pre-execute hook | `as_pre_execute_hook()` | securer integration |
| Output guard | `guard_output()` | Post-execution filtering |

Build small guards that each target one threat. Combine them with
`compose_guardrails()` or `check_all()`, and wire them into pipelines that
check every stage of an agent workflow. When the built-in guardrails do not
cover your domain, write your own with `new_guardrail()`. Custom guards and
built-in guards have the same interface and compose the same way.
